Probabilistic Catalogs for Crowded Stellar Fields

### 2012-10-24 paper/ms.tex
A few words, keywords and an equation fix

\begin{abstract}We introduce a probabilistic (Bayesian) method for producing catalogs
from images of crowded stellar fields. The method is capable of handling the
challenges introduced by overlapping sources and is able to determine the luminosity
function of the stars even when the precise luminosity of each star is uncertain.
This is in contrast with standard techniques which make a single catalog,
potentially underestimating the uncertainties in any study of the stellar
population. The method is implemented using advanced Markov Chain Monte Carlo
(MCMC) techniques including Reversible Jump and Nested Sampling.
The computational feasibility of the method is
demonstrated on simulated data where the luminosity function of the stars is a
broken power-law. The parameters of the luminosity function can be recovered
with moderate uncertainty.
\end{abstract}

@@

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data; it describes the probability distribution we would use
to predict the data if we happened to know the true catalog. In our case,
the data will be an $m \times m$ array of pixel intensities $I$:
\begin{eqnarray}
\{I_{ij}\}
\end{eqnarray}
where the central position of each pixel is:
\begin{eqnarray}
\{X_{ij}, Y_{ij}\}.
\end{eqnarray}
The image is assumed to be a noisy version of the true underlying intensity
field. Thus, we need
a prescription for simulating an image $\{I_{ij}\}$ from a catalog $\theta$:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N, 
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}
If we knew the true catalog, we could compute the
``mock image'' we would expect to see
in the absence of noise. This mock image (defined at every point on the sky)
is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved point spread function (PSF).
Throughout this paper we
assume the pixel-convolved PSF is a known weighted mixture of two concentric
circular
Gaussians with widths $s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus Gaussian noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma^2).
\end{eqnarray}
Throughout this paper, we will assume that the noise variance $\sigma^2$ is
known. This is not usually true in practice (although it is often assumed).
However it is straightforward to extend the model so that $\sigma$ is an
unknown parameter or that the noise variance should be a function of the
intensity of the mock image \citep[e.g.][]{2011MNRAS.412.2521B}. Allowing
the parameters of the pixel-convolved PSF to be unknown is another possible
extension. For the generation of the simulated data in Section~\ref{sec:simulated_data},
the values of the PSF parameters used were $s_1=0.02$, $s_2=0.1$ (one and
five pixel widths respectively), and $w=0.5$.

>>>>

\begin{abstract}We introduce a probabilistic (Bayesian) method for producing catalogs
from images of crowded stellar fields. The method is capable of handling the
challenges introduced by overlapping sources and is able to determine the luminosity
function of the stars even when the precise luminosity of each star is uncertain.
This is in contrast with standard techniques which make a single catalog,
potentially underestimating the uncertainties in any study of the stellar
population. The method is implemented using advanced Markov Chain Monte Carlo
(MCMC) techniques including Reversible Jump and Nested Sampling.
The computational feasibility of the method is
demonstrated on simulated data where the luminosity function of the stars is a
broken power-law. The parameters of the luminosity function can be recovered
with moderate uncertainty. We compare the results obtained from our method
with those obtained from the SExtractor software and find that the latter is
unable to correctly infer the number of stars or the luminosity function
parameters.
\end{abstract}

\keywords{catalogs --- methods: data analysis --- methods: statistical ---
stars: luminosity function, mass function}

@@

\subsection{The Sampling Distribution}
The sampling distribution is a probabilistic model for the process that
generates the data; it describes the probability distribution we would use
to predict the data if we happened to know the true catalog. In our case,
the data will be an $m \times m$ array of pixel intensities $I$:
\begin{eqnarray}
\{I_{ij}\}
\end{eqnarray}
where the central position of each pixel is:
\begin{eqnarray}
\{X_{ij}, Y_{ij}\}.
\end{eqnarray}
The image is assumed to be a noisy version of the true underlying intensity
field. Thus, we need
a prescription for simulating an image $\{I_{ij}\}$ from a catalog $\theta$:
\begin{eqnarray}
\theta = \left\{N, \beta, \left\{x_i, y_i\right\}_{i=1}^N, 
\left\{f_i\right\}_{i=1}^N\right\}.
\end{eqnarray}
If we knew the true catalog, we could compute the
``mock image'' we would expect to see
in the absence of noise. This mock image (defined at every point on the sky)
is given by:
\begin{eqnarray}
\mathcal{M}(x, y) &=& \sum_{i=1}^N f_i \mathcal{P}(x - x_i, y - y_i)
\end{eqnarray}
where $\mathcal{P}$ is the pixel-convolved point spread function (PSF).
Throughout this paper we
assume the pixel-convolved PSF is a known weighted mixture of two concentric
circular
Gaussians with widths $s_1$ and $s_2$:
\begin{eqnarray}
\mathcal{P}(x, y) = \frac{w}{2\pi s_1^2}\exp
\left[
-\frac{1}{2s_1^2}\left(x^2 + y^2\right)
\right]
+ \frac{1-w}{2\pi s_2^2}\exp
\left[-\frac{1}{2s_2^2}\left(x^2 + y^2\right)
\right].
\end{eqnarray}
The pixellated observed image is assumed to be generated from the mock
image (evaluated at the center of each pixel) plus Gaussian noise:
\begin{eqnarray}
I_{ij} &=& \mathcal{M}(X_{ij}, Y_{ij}) + \epsilon_{ij}
\end{eqnarray}
where the errors $\{\epsilon_{ij}\}$ are independent and normally distributed:
\begin{eqnarray}
\epsilon_{ij} \sim \mathcal{N}(0, \sigma^2).
\end{eqnarray}
Throughout this paper, we will assume that the noise variance $\sigma^2$ is
known. This is not usually true in practice (although it is often assumed).
However it is straightforward to extend the model so that $\sigma$ is an
unknown parameter or that the noise variance should be a function of the
intensity of the mock image \citep[e.g.][]{2011MNRAS.412.2521B}. Allowing
the parameters of the pixel-convolved PSF to be unknown is another possible
extension. For the generation of the simulated data in Section~\ref{sec:simulated_data},
the values of the PSF parameters used were $s_1=0.02$, $s_2=0.1$ (one and
five pixel widths respectively), and $w=0.5$.
